Natural selection in the Health and Retirement Study

### 2023-09-15 hrs-selection.qmd
Add siblings regressions

---
title: "Natural selection in the Health and Retirement Study"
author: "David Hugh-Jones"
date: "September 2023"
format: 
  pdf:
    geometry:
    - top=2in
    - left=1.5in
    - right=1.5in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

@@

@@

@@

for (eth in eths) {
  pgen <- pgens[[eth]]
  names(pgen) <- shorten(names(pgen))
  pgs <- pgen |>
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -EDU2, -EDU3, -HEIGHT, -BMI,
           -AD, -AD2, -GENCOG, -EVRSMK, -LONG, -CPD_TAG10,
           -ADHD_PGC10, MDD
           ) |>
    names()

@@

  rand <- rand |>
    mutate(
      mean_raevbrn = mean(raevbrn, na.rm = TRUE),
      rlrs = raevbrn/mean_raevbrn,
      .by = rabyear
    ) |>
    mutate(
      income_resid = calc_income(rand),
      income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
      income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
      # as.numeric avoids haven.labelled issues:
      raedyrs = as.numeric(raedyrs),
      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")) 
    )

@@

Hugh-Jones and Abdellaoui (2022) explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from Becker (XXX). The theory has two components.

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_edyrs <- mod_pgs_edyrs <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
cors_fert_edu <- tidy_pgs_joined <- list()

for (eth in eths) {
  rand <- rands[[eth]]

@@

  tidy_pgs[[eth]] <- list_rbind(map(
    mod_pgs[[eth]],
    broom::tidy,
    conf.int = TRUE,
    conf.level = 0.95
  )) |>
    filter(term != "(Intercept)", ! term %in% pcs)

@@

  tidy_pgs_edyrs[[eth]] <-
    list_rbind(map(mod_pgs_edyrs[[eth]], broom::tidy)) |>
    filter(term != "(Intercept)", ! term %in% pcs)

  tidy_pgs_joined[[eth]] <- inner_join(tidy_pgs[[eth]], tidy_pgs_edyrs[[eth]],
                                by = join_by(term),
                                unmatched = "error",
                                suffix = c(".fert", ".edyrs"),
                                relationship = "one-to-one")
  cors_fert_edu[[eth]] <- cor(tidy_pgs_joined[[eth]]$estimate.edyrs, 
                     tidy_pgs_joined[[eth]]$estimate.fert)

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) lm(x, data = rand, weights = weight)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) lm(x, data = rand, weights = weight)
        )

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "income_medHigh", pcs)) |>
    mutate(income_med = str_remove(term, ".*:income_med"))
}

@@

```{r}
#| label: fig-fertility
#| fig-cap: |
#|   Effects of PGS on realized fertility. Values are effects of a one
#|   standard deviation (within ethnicity) change of the PGS on fertility,
#|   controlling for 10 principal components of genomic array data. Scales
#|   differ between the two charts.
#| fig-height: 6

tidy_pgs_eth |>
  mutate(
    term = fct_reorder(term, estimate, \(x) x[2]),
    ethnicity = ifelse(ethnicity == "white", "Whites (non-Hispanic)", "Blacks"),
    # hack to order scales separately from 
    # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
    term_mangled = paste(ethnicity, term, sep = "/"),
    term_mangled = fct_reorder(term_mangled, estimate),
  ) |>
  ggplot(aes(estimate, term_mangled, color = ethnicity)) +
    geom_vline(xintercept = 0) +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                    position = position_dodge(width = 0.5), 
                    alpha = 0.8, size = 0.3) +
    facet_wrap(vars(ethnicity), scales = "free") +
    theme_light() + 
    scale_y_discrete(labels = \(x) str_remove(x, "^.*/")) +
    scale_color_manual(values = c("orange2", "seagreen")) +
    labs(
      x = "", y = "",
      title = "Effects of polygenic scores on fertility"
    ) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none",
      axis.text.y = element_text(size = 7),
      strip.text = element_text(colour = "black", size = 10),
      strip.background = element_rect(fill = NA)
    )
```

```{r}
#| label: stat-fertility-edu
#| cache: true

cors_boot <- list()
for (eth in eths) {
  rand_short <- rands[[eth]] |> 
    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight)
  cors_boot[[eth]] <-
    replicate(199, {
      rand <- rand_short |> 
              slice_sample(prop = 1, replace = TRUE)

@@

@@

      cor(coef_pgs_fert, coef_pgs_edyrs)
    })
}

tails_boot <- map(cors_boot, \(x) quantile(x, c(0.025, 0.975)))

if (! all(tails_boot$white < 0)) {
  stop("Fertility-edu correlation NOT sig at 95% for whites")
}
if (all(tails_boot$black < 0)) {
  stop("Fertility-edu correlation IS sig at 95% for blacks")
}

@@

@fig-fertility-edu plots each PGS's correlation with fertility against its correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white`, bootstrapped 95% C.I. `r tails_boot$white[1]` to `r tails_boot$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black`, bootstrapped 95% C.I. `r tails_boot$black[1]` to `r tails_boot$black[2]`). I use bootstraps to make inferences from the sample of individuals, rather than the "sample" of different PGS.

The relationship is imprecisely estimated in the smaller black sample, but it is also steeper. Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites. One reason for this may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites. In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf. @goldscheider1969minority; @johnson1979minority; @yang2003big]. In turn, the simplest explanation for this fact would be point 2 of the theory. Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them. TODO what does the existing literature say?

TODO if so then presumably controlling for education and income would reduce these differences.

```{r}
#| label: fig-fertility-edu
#| fig-cap: |
#|   Scatterplot of PGS partial correlations with fertility and years of
#|   education. Controls include 10 principal components of genetic array
#|   data. Dashed lines show linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

@@

>>>>

---
title: "Natural selection in the Health and Retirement Study"
author: "David Hugh-Jones"
abstract: |
  I investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  support the economic theory of fertility as an explanation for
  natural selection: among both white and black respondents,
  scores which correlate negatively (positively) with education are
  selected for (against). However, selection effects 
  are not larger among people with less education. 
  For whites, selection effects are smaller among married people. 
date: "September 2023"
format: 
  pdf:
    geometry:
    - top=2in
    - left=1.5in
    - right=1.5in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

# TODO: siblings E5669 "how many have died"
#   rands$white |> select(matches("r\\d+livsib")) -> tmp
#   apply(tmp, 1, max)

# TODO: use survey package
# here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd

# TODO: ratio to family poverty threshold, h10inpovr

# TODO: age at first birth. RAND family data provides h*ageokid
# TODO: RAND family data has r*sbdied for how many siblings have died

@@

nice_names <- c(
  SCZ         = "Schizophrenia",
  WC          = "Waist circ.",
  WHR         = "Waist-hip ratio",
  NEUROT      = "Neuroticism",
  WELLB       = "Well-being",
  DEPSYMP     = "Depr. symptoms",
  CD          = "Cor. art. disease",
  MI          = "Myoc. infarction",
  CORT        = "Cortisol",
  T2D         = "Type 2 diabetes",
  BIP         = "Bipolar",
  ADHD_PGC17  = "ADHD",
  XDISORDER   = "Cross-disorder",
  MENA        = "Age at menarche",
  MENO        = "Age at menopause",
  EXTRAVER    = "Extraversion",
  AUTISM      = "Autism",
  AB          = "Antisoc. behav.",
  OCD         = "OCD",
  AFBC        = "Age 1st birth combn",
  AFBF        = "Age 1st birth F",
  AFBM        = "Age 1st birth M",
  MDD2        = "Maj. depr. disorder",
  PTSDAA      = "PTSD Afr",
  PTSDEA      = "PTSD Eur",
  PTSDC       = "PTSD combined",
  HDL         = "HDL cholesterol",
  LDL         = "LDL cholesterol",
  TC          = "Total cholesterol",
  ANXFS       = "Anxiety CC", 
  ANXCC       = "Anxiety factor",
  BUN         = "Blood urea nitro.",
  BUNTE       = "Blood urea nitro. TE",
  CKD         = "Kidney disease",
  CKDTE       = "Kidney disease TE",
  DBP         = "Diastolic blood pr.",
  BMI2        = "BMI",
  HEIGHT2     = "Height",
  AI          = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW         = "Drinks per week",
  SC          = "Smoking cessation",
  SI          = "Smoking initiation",
  HTN         = "Hypertension",
  CANNABIS    = "Cannabis",
  GWAD2NA     = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA     = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC         = "Alcohol dependence",
  PP          = "Pulse pressure",
  SBP         = "Systolic blood pr.",
  EGFR        = "Glom. filtr.",
  EGFRTE      = "Glom. filtr. TE",
  EA3         = "Educ. attainment",
  HBA1CAA     = "HbA1c Afr",
  HBA1CEA     = "HbA1c Eur",
  GCOG2       = "Gen. cognition"
)

@@

# rand family respondents file
rand_fam <- haven::read_dta(
  "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)

@@

for (eth in eths) {
  pgen <- pgens[[eth]]
  names(pgen) <- shorten(names(pgen))
  pgs <- pgen |>
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -EDU2, -EDU3, -HEIGHT, -BMI,
           -AD, -AD2, -GENCOG, -EVRSMK, -LONG, -CPD_TAG10,
           -ADHD_PGC10, -MDD
           ) |>
    names()

@@

  rand <- rand |>
    mutate(.by = rabyear,
      mean_raevbrn = mean(raevbrn, na.rm = TRUE),
      rlrs = raevbrn/mean_raevbrn
    ) |>
    mutate(
      income_resid = calc_income(rand),
      income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
      income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
      # as.numeric avoids haven.labelled issues:
      raedyrs = as.numeric(raedyrs),
      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
      # wave 10 is 2010
      # table(rands$white$r10mstat)
       # 1                 1.married
       # 2   2.married,spouse absent
       # 3               3.partnered
       # 4               4.separated
       # 5                5.divorced
       # 6      6.separated/divorced
       # 7                 7.widowed
       # 8           8.never married
      married = factor(r10mstat == 1, 
                       levels = c(FALSE, TRUE),
                       labels = c("Other", "Married")),
    )

  rand <- rand |> left_join(rand_fam, 
                            by = join_by(hhid, pn),
                            unmatched = "drop",
                            relationship = "one-to-one")
  rand <- rand |> mutate(
    r10sibs = r10livsib + r10sbdied
  )

@@

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
for (eth in eths) {
  rand <- rands[[eth]]

@@

  tidy_pgs[[eth]] <- mod_pgs[[eth]] |> 
    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
    list_rbind() |>
    filter(term %in% pgs)

@@

  tidy_pgs_ed[[eth]] <- mod_pgs_edyrs[[eth]] |>  
    map(broom::tidy) |>
    list_rbind() |> 
    filter(term %in% pgs)

  tidy_pgs_joined[[eth]] <- inner_join(tidy_pgs[[eth]], tidy_pgs_ed[[eth]],
                                by = join_by(term),
                                unmatched = "error",
                                suffix = c(".fert", ".edyrs"),
                                relationship = "one-to-one")
  cors_fert_edu[[eth]] <- cor(tidy_pgs_joined[[eth]]$estimate.edyrs, 
                     tidy_pgs_joined[[eth]]$estimate.fert)

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                             \(x) lm(x, data = rand, 
                                     weights = weight/(1 + r10sibs)))

  tidy_pgs_sibs[[eth]] <- mod_pgs_sibs[[eth]] |> 
    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |>
    list_rbind() |> 
    filter(term %in% pgs)

  cors_sibs_edu[[eth]] <- cor(tidy_pgs_sibs[[eth]]$estimate, 
                     tidy_pgs_ed[[eth]]$estimate)

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) lm(x, data = rand, weights = weight)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) lm(x, data = rand, weights = weight)
        )

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "income_medHigh", pcs)) |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
        \(x) lm(x, data = rands[[eth]], weights = weight)
        )

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
    mutate(
      married = str_remove(term, ".*:married")
    ) 
}

@@

```{r}
#| label: fig-fertility
#| fig-cap: |
#|  Effects of PGS on realized fertility. Values are partial correlations 
#|  of a one standard deviation (within ethnicity) change of the PGS with
#|  fertility, controlling for 10 principal components of genomic array 
#|  data. Scales differ between the two charts.
#| fig-height: 6

tidy_pgs_eth |>
  mutate(
    term = nice_names[term],
    term = fct_reorder(term, estimate, \(x) x[2]),
    ethnicity = ifelse(ethnicity == "white", "Whites (non-Hispanic)", "Blacks"),
    # hack to order scales separately from 
    # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
    term_mangled = paste(ethnicity, term, sep = "//"),
    term_mangled = fct_reorder(term_mangled, estimate),
  ) |>
  ggplot(aes(estimate, term_mangled, color = ethnicity)) +
    geom_vline(xintercept = 0) +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                    position = position_dodge(width = 0.5), 
                    alpha = 0.8, size = 0.3) +
    facet_wrap(vars(ethnicity), scales = "free") +
    theme_light() + 
    scale_y_discrete(labels = \(x) str_remove(x, "^.*//")) +
    scale_color_manual(values = c("orange2", "seagreen")) +
    labs(
      x = "", y = "",
      title = "Effects of polygenic scores on fertility"
    ) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none",
      axis.text.y = element_text(size = 7),
      strip.text = element_text(colour = "black", size = 10),
      strip.background = element_rect(fill = NA)
    )
```

```{r}
#| label: stat-bootstraps
#| cache: true

n_reps <- 599
cors_boot <- list()
for (eth in eths) {
  rand_short <- rands[[eth]] |> 
    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, live_sibs)
  cors_boot[[eth]] <-
    replicate(n_reps, {
      rand <- rand_short |> 
              slice_sample(prop = 1, replace = TRUE)

@@

      coef_pgs_sibs <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "live_sibs")) |> 
        map(\(x) lm(x, data = rand, weights = weight/(live_sibs + 1))) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

      c(fert_edyrs = cor(coef_pgs_fert, coef_pgs_edyrs),
        sibs_edyrs = cor(coef_pgs_sibs, coef_pgs_edyrs))
    })
}

tails_boot_fert <- map(cors_boot, 
                       \(x) quantile(x["fert_edyrs", ], c(0.025, 0.975)))

tails_boot_sibs <- map(cors_boot,
                       \(x) quantile(x["sibs_edyrs", ], c(0.025, 0.975)))

if (! all(tails_boot_fert$white < 0)) {
  stop("Fertility-edu correlation NOT sig at 95% for whites")
}
if (all(tails_boot_fert$black < 0)) {
  stop("Fertility-edu correlation IS sig at 95% for blacks")
}

@@

@fig-fertility-edu plots each PGS's correlation with fertility against its correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white`, bootstrapped 95% C.I. `r tails_boot_fert$white[1]` to `r tails_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black`, bootstrapped 95% C.I. `r tails_boot_fert$black[1]` to `r tails_boot_fert$black[2]`). I use bootstraps so as to make inferences from the sample of individuals, rather than the "sample" of different PGS.

The relationship is imprecisely estimated in the smaller black sample, but it is also steeper. Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites. If this holds in the population, one reason may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites. In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf. @goldscheider1969minority; @johnson1979minority; @yang2003big]. In turn, the simplest explanation for this fact would be point 2 of the theory. Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them.

```{r}
#| label: fig-fertility-edu
#| fig-cap: |
#|   Scatterplot of PGS effect sizes on fertility and years of
#|   education. Controls include 10 principal components of genetic array
#|   data. Dashed lines show linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

@@

```{r}
#| label: stat-siblings

tmp <- 
  tidy_pgs_sibs$black |> 
  inner_join(tidy_pgs_ed$black, 
             by = join_by(term), 
             suffix = c(".sibs", ".edyrs")) |> 
  filter(term != "GWAD2NA", term != "GWAD2WA")

cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
```

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010. I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so effect sizes are not comparable across the generations. @fig-siblings plots effects on number of siblings versus effects on years of education. The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I. `r tails_boot_sibs$white[1]` to `r tails_boot_sibs$white[2]`). It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.

```{r}
#| label: fig-siblings
#| fig-cap: |
#|   Scatterplot of PGS effect sizes on number of live siblings and years of
#|   education. Controls include 10 principal components of genetic array
#|   data. Dashed lines show linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

for (eth in eths) {
  ggp <-  
    tidy_pgs_sibs[[eth]] |> 
    inner_join(tidy_pgs_ed[[eth]], 
               by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
    ggplot(aes(estimate.edyrs, estimate.sibs)) +
      geom_hline(yintercept = 0, color = "grey80") +
      geom_vline(xintercept = 0, color = "grey80") +
      geom_point() +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                  linetype = "dashed", colour = "red", linewidth = 0.5) +
      coord_cartesian(xlim = c(-.3, .75)) +
      theme_light() + 
      theme(panel.grid = element_blank()) +
      labs(
          title = eth,
          x = "Effect on years of education",
          y = "Effect on number of siblings"
      )
  print(ggp)
}

I next test part 2 of the theory more directly by interacting PGS with measures of education, income and marital status. Education is years of education, split at 12 years. Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split.

```{r}
#| label: stat-x-edu
#| cache: true

n_reps <- 199

cross_terms <- function (x) {
  cx <- coef(x)
  cx[str_detect(names(cx), ":")]
}

form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
form_pgs_x_inc <- paste0(pgs, ":income_med")  |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
form_pgs_x_mar <- paste0(pgs, ":married")  |>
    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))

coefs_boot_ed <- coefs_boot_inc <- coefs_boot_mar <- list()

for (eth in eths) {
  rand <- rands[[eth]]
  rand_short <- rands[[eth]] |> 
    select(
      all_of(pgs), all_of(pcs), 
      raevbrn, raedyrs, weight, income_med, edu, married
    )

  coefs_boot_ed[[eth]] <- replicate(n_reps, {
    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
    map(form_pgs_x_ed, 
        \(x) lm(x, data = rand_r, weights = weight)
        ) |> 
      map(cross_terms) |> 
      unlist()
  })

  coefs_boot_inc[[eth]]  <- replicate(n_reps, {
    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
    map(form_pgs_x_inc, 
          \(x) lm(x, data = rand_r, weights = weight)
          ) |> 
      map(cross_terms) |> 
      unlist()
  })

  coefs_boot_mar[[eth]]  <- replicate(n_reps, {
    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
    map(form_pgs_x_mar, 
          \(x) lm(x, data = rand_r, weights = weight)
          ) |> 
      map(cross_terms) |> 
      unlist()
  })

  reorganize <- function (x) {
   x |> 
    t() |> 
    as.data.frame() |> 
    pivot_longer(everything(), names_to = c("pgs", "group"), 
                 names_sep = ":") |> 
    mutate(rep = 1:n_reps, .by = c(pgs, group)) |> 
    pivot_wider(id_cols = c(pgs, rep), 
                names_from = group, 
                values_from = value) 
  }

  coefs_boot_ed[[eth]] <- reorganize(coefs_boot_ed[[eth]])
  coefs_boot_inc[[eth]] <- reorganize(coefs_boot_inc[[eth]])
  coefs_boot_mar[[eth]] <- reorganize(coefs_boot_mar[[eth]])
}

boot_ci <- function (x) {
  names(x)[3:4] <- c("lo", "hi")
  x |> 
    mutate(
      abs_diff = abs(hi) - abs(lo)
    ) |> 
    summarize(
      mad = mean(abs_diff),
      .by = rep
    ) |> 
    pull(mad) |> 
    quantile(c(0.025, 0.975))
}

ci_inc_w <- boot_ci(coefs_boot_inc$white)
ci_inc_b <- boot_ci(coefs_boot_inc$black)
ci_ed_w <- boot_ci(coefs_boot_ed$white)
ci_ed_b <- boot_ci(coefs_boot_ed$black)
ci_mar_w <- boot_ci(coefs_boot_mar$white)
ci_mar_b <- boot_ci(coefs_boot_mar$black)

for (ci in list(ci_inc_w, ci_inc_b, ci_ed_w, ci_ed_b)) {
  if (prod(sign(ci)) != -1) {
    stop("effect size IS significantly different across some pair groups")
  }  
}
```

@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education. @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks. However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group. @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses. Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents, and this is significant in bootstraps (95% C.I. for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`).

```{r}
#| label: fig-x-edu
#| fig-cap: |
#|   Violin plots of polygenic score effects on fertility by education. 
#|   Effect sizes are
#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
#|   change of the PGS on fertility, controlling for 10 principal 
#|   components of genetic array data, estimated within the low/high
#|   education group. Lines show quartiles.

tidy_pgs_x_ed |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = unlist(titles[ethnicity]),
    Education = str_remove(edu, "edu")
  ) |> 
  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Years of education",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

```{r}
#| label: fig-x-inc
#| fig-cap: |
#|   Violin plots of polygenic score effects on fertility by income. 
#|   Effect sizes are
#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
#|   change of the PGS on fertility, controlling for 10 principal 
#|   components of genetic array data, estimated within the low/high income
#|   group. Lines show quartiles.

tidy_pgs_x_inc |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = unlist(titles[ethnicity]),
    Income = fct_relevel(income_med, c("Low", "High"))
  ) |> 
  ggplot(aes(Income, abs(estimate), color = Income, fill = Income)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Income",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

```{r}
#| label: fig-x-mar
#| fig-cap: |
#|   Violin plots of polygenic score effects on fertility by marital status. 
#|   Effect sizes are
#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
#|   change of the PGS on fertility, controlling for 10 principal 
#|   components of genetic array data, estimated within the low/high income
#|   group. Lines show quartiles.

tidy_pgs_x_mar |> 
    list_rbind(names_to = "ethnicity") |> 
    mutate(
        Ethnicity = unlist(titles[ethnicity]),
        married = fct_relevel(married, "Other", "Married")
    ) |> 
    ggplot(aes(married, abs(estimate), color = married, fill = married)) +
      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
      facet_wrap(vars(Ethnicity), scales = "free_y") +
      labs(
          x = "Married",
          y = "Absolute effect size"
      ) +
      theme_minimal() +
      theme(
          legend.position = "none"
      )

# Discussion

The results here provide qualified support for the economic theory of fertility as an explanation for contemporary natural selection in humans. PGS which predict less education are being selected for, and PGS which predict more education are being selected against. However, there is no evidence that PGS effects on fertility are larger among people with less education, unlike the UK [@hugh2022human]. And there's not enough evidence to show that effects on fertility are larger for lower-income people. Also, the smaller black sample made most tests inconclusive for this population.
